---
# MB5370: Introduction to Programming
title: "Wrangling_Noisy_Data_M2W2"
author: "Ryan Waln"
format: html
editor: visual
---

## Introduction

This file is used to wrangle messy data on Ross River estuary fish surveys and produce final clean tables and figures to determine if fish species distributions are related to salinity

#### Housekeeping

```{r}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

#### Packages

```{r}
library("tidyverse")
library("readxl") 
```

## Read in data 

```{r}
estuary_metadata   <- read_csv(here::here("data", "workshop_2", "estuary_metadata.csv"))
estuary_sonde_data <- read_csv(here::here("data", "workshop_2","estuary_sonde_data.csv"))
species_dictionary <- read_csv(here::here("data", "workshop_2","species_dictionary.csv"))
```

#### Check site locations

```{r}
estuary_metadata
head(estuary_metadata)
glimpse(estuary_metadata)
str(estuary_metadata)
```

#### Check sensor data (dates are in character format, legacy errors, and not 

```{r}
estuary_sonde_data
head(estuary_sonde_data)
glimpse(estuary_sonde_data)
str(estuary_sonde_data)
```

#### Check species dictionary

```{r}
species_dictionary
head(species_dictionary)
glimpse(species_dictionary)
str(species_dictionary)
```


#### Combine excel tabs on catch log into one data frame so all sheets load

```{r}
catch1 <- read_excel(here::here("data/workshop_2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment

# Glimpse data
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#### Alternative method for loading files and combining sheets

```{r}
file_path <- "../data/workshop_2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |>
  map_dfr(~read_excel(path = file_path, sheet = .x))


catch_data2
head(catch_data2)
glimpse(catch_data2)
str(catch_data2)
```

#### Check Species and Site Names (no scientific names & inconsistent syntax)

```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

## Data Cleaning

#### Standardize text with String

```{r}

#Standardize species and site names on catch log
catch_log <- catch_data |>
  mutate(
    # 1. Convert everything to lowercase
    species = str_to_lower(species),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    species = str_trim(species),
    # 3. Replace any spaces with underscores
    species = str_replace_all(species, pattern = " ", replacement = "_")
  ) |>
mutate(
    # 1. Convert everything to lowercase
    site = str_to_lower(site),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site = str_trim(site),
    # 3. Replace any spaces with underscores
    site = str_replace_all(site, pattern = " ", replacement = "_")
  )

print(catch_log)


#Standardize common names on species dictionary to join with species on catch data later
#DONT NEED
species_dictionary <- species_dictionary |>
  mutate(
    # 1. Convert everything to lowercase and replace old column 
    common_name = str_to_lower(common_name),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    common_name = str_trim(common_name),
    # 3. Replace any spaces with underscores
    common_name = str_replace_all(common_name, pattern = " ", replacement = "_")
  )

print(species_dictionary)


# Standardize site names on metadata to join with site on catch log later 
estuary_metadata <- estuary_metadata |>
  mutate(
    # 1. Convert everything to lowercase and replace old column 
    site_name = str_to_lower(site_name),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_name = str_trim(site_name),
    # 3. Replace any spaces with underscores
    site_name = str_replace_all(site_name, pattern = " ", replacement = "_")
  )

print(estuary_metadata)
```

#### Add scientific names by merging 
```{r}

clean_catch_log <- catch_log |>
  left_join(species_dictionary, by = join_by(species == common_name))  #ensure variable on left is from dataset specified on line above

print(clean_catch_log)

```

#### Formate dates and NA values in Sonde Dataset 

```{r}
clean_sensor_data = estuary_sonde_data |>
  mutate( 
    # change timestamp to date_time format
    timestamp = dmy_hm(timestamp),
    # covert hardware error to NA values
    turbidity = na_if(turbidity, -999.0),
    # convert negative salinity values to NA
    salinity = ifelse(salinity < 0, NA, salinity)
    )
```

## Put Data into Cohesive Format

#### Put sensor data in daily format
```{r}
# Sensor data is logged every 15 minutes, but catch data is daily. Create a daily summary table that calculates the mean temperature and salinity for each site on each day. 

daily_sensor <- clean_sensor_data |>
  mutate(
    # Round each timestamp DOWN to midnight of that day
    date = floor_date(timestamp, unit = "day"),
    # Ensure it matches date format of catch log by converting timestamp 
    date = as_date(date)
  ) |>
  # Group by site and date to calculate for each  unique combination of the two
  group_by(site, date) |>
  # Calculate means and remove NA values
  summarise(
    mean_temp     = mean(temperature, na.rm = TRUE),
    mean_salinity = mean(salinity,    na.rm = TRUE),
    # Remove grouping so table displays normally
    .groups = "drop"
  )

daily_sensor
```

#### Merge catch data, spatial metadata, and daily water quality summaries into a single master data frame.

```{r}
# Merge clean_catch_log, estuary_metadata, and your daily_sensor into a single master data frame

final_catch_data <- clean_catch_log |>
  left_join(estuary_metadata, by = join_by(site == site_name)) |>
  left_join(daily_sensor, by = join_by(site, date))

final_catch_data
```

#### Add rows for when no individuals of each species were caught to ensure that every species is represented at every site, on every sampling day.



```{r}
final_catch_data |> 
  complete(
    nesting(site, date, 
           # lat, lon, 
            zone, mean_temp, mean_salinity),
    scientific_name
  ) |>
  mutate(
    # Convert the resulting NA counts to 0
    count = coalesce(count, 0)
  )
```

## Create summary table of salinity and fish counts with SE

```{r}
# From final_catch_data, generate a clean summary table showing the mean ± standard error of both count and the salinity for each scientific name across each  zone 

summary_table <- final_catch_data |>
  group_by(scientific_name, zone) |>
  summarise(
    # Calculate means and SE
    mean_count    = mean(count, na.rm = TRUE),
    se_count      = sd(count, na.rm = TRUE) / sqrt(n()),
    mean_salinity = mean(mean_salinity, na.rm = TRUE),
    se_salinity   = sd(mean_salinity, na.rm = TRUE) / sqrt(sum(!is.na(mean_salinity))),
    # Drop groups to present table
    .groups = "drop"
  )

print(summary_table)


```

## Plot Data

```{r}
# The plot should have professional axis labels, a clean theme (e.g., theme_minimal()), be ordered spatially (e.g. from upstream to downstream), and the facet labels (the scientific names) should really be italicised

summary_table <- summary_table |>
  mutate(zone = factor(zone, levels = c("upstream", "mid", "downstream")))

ggplot(summary_table, aes(x = mean_salinity, y = mean_count, colour = zone)) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_count - se_count,
                    ymax = mean_count + se_count),
                width = 0.3) +
  geom_errorbar (aes(xmin = mean_salinity - se_salinity,
                     xmax = mean_salinity + se_salinity),
                 height = 0.3) +
  facet_wrap(~ scientific_name,
             scales = "free_y",
             labeller = label_parsed) +
  theme_minimal() +
  labs(
    x      = "Mean salinity (ppt)",
    y      = "Mean catch (individuals)",
    colour = "Zone",
    title  = "Fish catch vs. salinity by species and zone"
  ) +
  theme(
    strip.text = element_text(face = "italic"),
    legend.position = "bottom"
  )
```

