---
# MB5370: Introduction to Programming
title: "Wrangling_Noisy_Data_M2W2"
author: "Ryan Waln"
format: html
editor: visual
---

## Introduction

This file is used to wrangle messy data on Ross River estuary fish surveys and produce final clean tables and figures

#### Housekeeping

```{r}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

#### Packages

```{r}
library("tidyverse")
library("readxl") 
library("dplyr")
```

## Read in data 

```{r}
estuary_metadata   <- read_csv(here::here("data", "workshop_2", "estuary_metadata.csv"))
estuary_sonde_data <- read_csv(here::here("data", "workshop_2","estuary_sonde_data.csv"))
species_dictionary <- read_csv(here::here("data", "workshop_2","species_dictionary.csv"))
```

#### Check site locations

```{r}
estuary_metadata
head(estuary_metadata)
glimpse(estuary_metadata)
str(estuary_metadata)
```

#### Check sensor data

```{r}
estuary_sonde_data
head(estuary_sonde_data)
glimpse(estuary_sonde_data)
str(estuary_sonde_data)
```

#### Check species dictionary

```{r}
species_dictionary
head(species_dictionary)
glimpse(species_dictionary)
str(species_dictionary)
```


#### Combine excel tabs on catch log into one data frame so all sheets load

```{r}
catch1 <- read_excel(here::here("data/workshop_2/estuary_catch_log.xlsx"), sheet = 1)
catch2 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop_2/estuary_catch_log.xlsx", sheet = 4)

catch_data <- bind_rows(catch1, catch2, catch3, catch4)
# Clean up the environment
rm(catch1, catch2, catch3, catch4) # Clean up the environment

# Glimpse data
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

#### Alternative method for loading files and combining sheets

```{r}
file_path <- "../data/workshop_2/estuary_catch_log.xlsx"
catch_data2 <- excel_sheets(file_path) |>
  map_dfr(~read_excel(path = file_path, sheet = .x))


catch_data2
head(catch_data2)
glimpse(catch_data2)
str(catch_data2)
```

#### Check Species and Site Names (no scientific names & inconsistent syntax)

```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```


## Data Cleaning

#### Change common names to scientific names

```{r}

```

#### Standardize text with String

```{r}
clean_catch_log <- catch_data |>
  mutate(
    # 1. Convert everything to lowercase
    species = str_to_lower(species),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    species = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    species = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)
```

#### Formate dates in Sonde Dataset 

```{r}

```

#### Convert legacy errors to NA values

```{r}

```

## Put Data into Cohesive Format

```{r}

```




