---
title: "R for Marine Science Workshop 1"
author: "Ryan Waln"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

## Introduction

This file is paired with a base R file to allow easy plotting of penguin data. 

Material covered on base R script includes:

- importing data 
- formatting data for R 
- Tidyverse tibbles vs base R data frames
- piping 
- mutating data 
- basic NA value handling

Material covered on this qmd file includes: 

- piping data to create plots
- plotting graphs with error bars
- saving tables and figures

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
#  R Setup & housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
rm(list=ls())
objects()
```

#### Load packages

```{r packages}
library(tidyverse)
```

#### Source script

```{r source-script1}
#| results: hide
source("Workshop1.R")
```

#### Pipe data to summarize mean body mass for each species and island and create a box plot of body mass by species, with the island variable mapped to fill

```{r}
mass_metrics = penguins %>%
  group_by(species, island) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(mass_metrics) 

penguins |>
  ggplot(aes(x = species, y = body_mass_g, fill = island)) +  #Plots with assigned x and y axis with islands as fill
  geom_boxplot(na.rm = TRUE) + # Stops missing value trap
  
  #Add labels and theme
  labs(
    title = "Penguin Body Mass by Species and Island",
    x = "Species",
    y = "Body Mass (g)",
    fill = "Island"
  )  +
  theme_minimal()
```

#### Instead of creating a summary table object first, you can pipe your grouped data straight into a plot and use geom_errorbar() to represent the spread of your data

```{r}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot = penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

#### Plot with standard error instead (always include table when presenting to others)

```{r}
mass_compare_plot2 = penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    se_mass = (sd_mass/ sqrt(n)), #calculate SE
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - se_mass, 
                    ymax = mean_mass + se_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard error",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot2

#Display summary table to help determine patterns
mass_metrics = penguins %>%
  group_by(species, island) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(mass_metrics)  
```

# Data suggests Gentoo Penguins are much bigger than other species and Adelie penguins do not vary much by mass across the islands


## Saving Work

### Can save Tables as:

#### Universal CSV text file: Ideal for sharing data tables with collaborators who do not use R.

#### Native R Binary RDS object: Saving as an RDS file preserves your exact vector configurations (such as custom categorical factor levels and numeric type definitions), meaning you never have to re-format column types when you reload the data next time you work on it.

### Save Tables

```{r}
# add ../ to go up one level since .Rmd & .qmd will always default to saving in exact loaction

# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "../outputs/penguin_species_mass_summary.csv")

# 2. Saving our cleaned morphological cohort table as a native R binary file
saveRDS(clean_scientific_fields, "../outputs/clean_penguin_morphology_cohort.rds")
```

### Save Plot

```{r}
ggsave("../outputs/figures/mass_compare_plot2.png", 
       plot = mass_compare_plot2, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)

```
