---
# MB5370: Introduction to Programming
title: "Wrangling_Plotting_M2W1"
author: "Ryan Waln"
date: "2026-06-02"
output: html_document
---

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Creating a Standalone Clone of a Repository
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

## Github setup (run in console)
```{r}
##Enter into terminal if using USB
#git config --global --add safe.directory "D:/Tri 2 2026/Techniques in Marine Science 1/Data_Science_in_R/R4MarineScience"

##Then restart R and run 
usethis::git_sitrep() 

##Get git credentials
credentials::git_credential_ask()  

## see Git status
usethis::git_sitrep()

usethis::use_git_config(
  user.name = "Ryan Waln",
  user.email = "ryan.waln@aol.com"
)

## create Personal Access Token
usethis::create_github_token() 

## prompt to insert PAT
#gitcreds::gitcreds_set() 

gert::git_remote_remove("origin") #severs connection to original repository to allow storage in own GitHub after cloning

##track current project locally (can also use Tools > Project Options … > Git/SVN, and set to Git)
## or run  git init in terminal
usethis::use_git() 

usethis::use_github() # Have github start tracking files
```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Creating a Standalone Clone of a Repository
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

## Housekeeping
```{r}
## Inventory every active object currently residing in session RAM
objects()
## Purge global environment
rm(list = ls())
# Confirm that global session memory now completely vacant
objects()

#In global options: Uncheck the box that says: Re-restore .RData into workspace at startup and set to Never

unlink("~/.RData") #Prevent old hanging code and values from screwing up current code
```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Packages
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

```{r}
#install.packages("gert")
#install.packages("usethis")
#install.packages("here")
#install.packages("tidyverse")
#install.packages("readxl")
```

```{r}
library("gert")
library("usethis")
library("here")
library("tidyverse")
library("readxl")
```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Importing Data
-----------------------------------------------------------------------------------------------------------------------------------------------------------------
 
## Importing Different File types
```{r}
# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/fish_catch_data.xlsx"), sheet = "Commercial_2026")

```

## Data not formated for R:
```{r}
# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/mangrove_survey_raw.csv"))

```

## Format data
```{r}
# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA

```

##Tidyverse Tibble vs base R Datframes (tibble lets you spot errors)
```{r}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)
# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)

```
-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Wrangling out ecological signals using Palmer Penguins dataset
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

```{r}
#install.packages("palmerpenguins")
library(palmerpenguins)
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

```

## Types of varriables
### <fct> (Factor): Categorical groupings with fixed levels (e.g., species containing Adelie, Chinstrap, and Gentoo).
### <dbl> (Double): Continuous numeric measurements containing decimals (e.g., bill_length_mm).
### <int> (Integer): Whole number variables, usually to track counts, (e.g., body_mass_g).

```{r}
# Generate an exploratory summary matrix
summary(penguins)

```

```{r}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)


```
-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# The Pipe |> or %>% (used to prevent clutter in Enviornment tab by linking functions together)
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

## Without a Pipe: You have to read the code "inside-out." You start in the middle, perform the mutate(), then wrap that in a filter(), and finally wrap that all in a select().

## With a Pipe: You read from left to right, or top to bottom. You take your data, then you filter it, then you mutate it to create a new column, then you select the columns you need.

## Example Syntax 

### Typical 
```{r}
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")
```

### Pipe
```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Mutating Data
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

## Mutating is how we modify data
```{r}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)

```

# Missing Value Trap: NA values stop functions from working
```{r}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)

```

```{r}
# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal) #penguins now grouped by species and sex for mean mass

```

-----------------------------------------------------------------------------------------------------------------------------------------------------------------
# Plotting
-----------------------------------------------------------------------------------------------------------------------------------------------------------------

## Grammer of Graphics

### The Data Layer: Declaring the source dataset table object.
### The Aesthetic Mapping (aes): Defining which variables are mapped to structural axes, colors, shapes, or sizes.
### The Geometric Layer (geom_...): Defining the visual shape that represents the numbers (e.g., points, bars, lines, or boxplots).


# After sorting data in basic R script (this markdown is for testing purposes), use a .qmd file to render work into a professional HTML document

