---
# MB5370: Data Science in R
title: "Wrangling_Noisy_Data_M2W2"
author: "Ryan Waln"
format: html
editor: visual
---

## Introduction

This file covers more advanced data wrangling techniques, including:

- Pivoting data to move around rows and columns in tables
- Separating and uniting tables or specific columns
- Standardizing text and dates in data sets
- Dealing with NA values

```{r}
# R Setup & housekeeping
knitr::opts_chunk$set(
	echo = FALSE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)

rm(list=ls())
objects()
```

```{r}
# Packages
library(tidyverse)
```

## Rules for tidy data:

- Each variable must have its own column.

- Each observation must have its own row.

- Each value must have its own cell.

- ALWAYS: Put each dataset in a tibble and each variable in a column.

#### Example of working with tidy data

```{r}
# Compute rate per 10,000 (existing data set)
table1 %>% 
  mutate(rate = cases / population * 10000)
#> # A tibble: 6 × 5
#>   country      year  cases population  rate
#>   <chr>       <int>  <int>      <int> <dbl>
#> 1 Afghanistan  1999    745   19987071 0.373
#> 2 Afghanistan  2000   2666   20595360 1.29 
#> 3 Brazil       1999  37737  172006362 2.19 
#> 4 Brazil       2000  80488  174504898 5.61 
#> 5 China        1999 212258 1272915272 1.67 
#> 6 China        2000 213766 1280428583 1.67

# Compute cases per year
table1 %>% 
  count(year, wt = cases)
#> # A tibble: 2 × 2
#>    year      n
#>   <int>  <int>
#> 1  1999 250740
#> 2  2000 296920

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```

## Pivoting data: Lengthening data sets

```{r}
# Usually used to fix when one variable is spread across multiple columns or one observation across multiple rows

# Increases rows & decreases columns:  pivot_longer() 

#Data on ranked songs of 2000 (each song's rank listed over 76 weeks)
billboard
#> # A tibble: 317 × 79
#>   artist       track               date.entered   wk1   wk2   wk3   wk4   wk5
#>   <chr>        <chr>               <date>       <dbl> <dbl> <dbl> <dbl> <dbl>
#> 1 2 Pac        Baby Don't Cry (Ke... 2000-02-26      87    82    72    77    87
#> 2 2Ge+her      The Hardest Part O... 2000-09-02      91    87    92    NA    NA
#> 3 3 Doors Down Kryptonite          2000-04-08      81    70    68    67    66
#> 4 3 Doors Down Loser               2000-10-21      76    76    72    69    67
#> 5 504 Boyz     Wobble Wobble       2000-04-15      57    34    25    17    17
#> 6 98^0         Give Me Just One N... 2000-08-19      51    39    34    26    26
#> # ℹ 311 more rows
#> # ℹ 71 more variables: wk6 <dbl>, wk7 <dbl>, wk8 <dbl>, wk9 <dbl>, ...

# Move weeks into one column to tidy
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), # specifies which columns to pivot
    names_to = "week",        # names the variable stored in the column names
    values_to = "rank",       # names the variable stored in the cell values that we named rank (had to create those 2 variables)
    values_drop_na = TRUE
  )
#> # A tibble: 24,092 × 5
#>    artist track                   date.entered week   rank
#>    <chr>  <chr>                   <date>       <chr> <dbl>
#>  1 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk1      87
#>  2 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk2      82
#>  3 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk3      72
#>  4 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk4      77
#>  5 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk5      87
#>  6 2 Pac  Baby Don't Cry (Keep... 2000-02-26   wk6      94
#>  #> # ℹ 24,082 more rows



```

#### Pivoting: longer example

```{r}
# Create dataset "df" w/3 varriables
df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

# Tidy data to have 3 variables: id (already exists), measurement (the column names), & value (the cell values)
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",  
    values_to = "value"    
  )


```

## Pivoting data: widening data sets

```{r}
cms_patient_experience  #load buit in data set expressing patient satisfaction w/ health organizations
#> # A tibble: 500 × 5
#>   org_pac_id org_nm                     measure_cd   measure_title   prf_rate
#>   <chr>      <chr>                      <chr>        <chr>              <dbl>
#> 1 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_1  CAHPS for MIPS...       63
#> 2 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_2  CAHPS for MIPS...       87
#> 3 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_3  CAHPS for MIPS...       86
#> 4 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_5  CAHPS for MIPS...       57
#> 5 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_8  CAHPS for MIPS...       85
#> 6 0446157747 USC CARE MEDICAL GROUP INC CAHPS_GRP_12 CAHPS for MIPS...       24
#> # ℹ 494 more rows

cms_patient_experience |> 
  distinct(measure_cd, measure_title)   #See complete set of variables
# measure_cd doesn’t hint at the meaning of the variable and measure_title is a long sentence containing spaces
```

Code below still results in multiple rows for each organization since we still need to tell pivot_wider() which column or columns have values that uniquely identify each row

```{r}
# add line to specify organizations
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

```

#### Widening data sets example:

```{r}
# Create data set with blood pressure measurments from 2 patients
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

# Rearrange data with patients as rows
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )
```

## Pivoting Data: Palmer Penguins

#### Lengthening data

```{r}
library(palmerpenguins)
head(penguins)

# Create a long version of the penguins data set

penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
# View the result
head(penguins_long)



#Create histogram of long data
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )


```

#### Widening data (can use to present summary statistics as a cross-tabulated matrix. It is much easier for a reader to parse a wide table than a long, repetitive list).

```{r}
# Calculate mean body mass for each species on each island 
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))

head(mass_summary)

# Output is technically "tidy" and computationally useful, but it is not ideal for publication presentation. We can use pivot_wider() to pull the island names up into columns, creating a clean matrix of species by island.
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )
head(mass_matrix)

# names_from dictates which column provides our new column headers (the islands), and values_from dictates which column populates the cells beneath them (the mean mass)
```

## Separating and uniting data tables

```{r}
# Table where cases and population are combined into one column to show rate of infection per 10,000 people
table3

# We need to split the rate column up into two variables: 1) cases and 2) population
table3 %>% 
  separate(rate, into = c("cases", "population"))

# Can also specify what character to use when separating:
# separate(rate, into = c("cases", "population"), sep = "/")

# Need to use convert = TRUE to set new variables as numerical since separate() defaults to character
table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

# Can also separate numerical varriables by position instead of character (less useful)
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)

# Uniting columns is also an option
table5
table5 %>% 
  unite(new, century, year, sep = "")  #unites century and year into column called "new" without a separator

```

## Wrangling Strings and Dates

Before we can merge datasets or plot timelines, we must standardize our text strings and date-times. The tidyverse includes two specialized packages for this: stringr for text and lubridate for dates.

#### Using Stringr to standardize text

```{r}
library(tidyverse)

# A remarkably messy data frame of field sites
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)

# Using stringr within mutate to standardize the text
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
  )

print(clean_sites)


```

#### Using Lubridate to standardize Dates

```{r}
library(lubridate)

# Parsing different date formats
date_1 <- dmy("25/12/2026") # if day is listed first
date_2 <- ymd("2026-12-25") # if year is listed first

# R now recognizes these as identical Date objects
date_1 == date_2

# A tibble of raw sensor data with a messy character timestamp
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)

# Converting character strings to true POSIXct datetime objects
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )

print(sensor_clean)
```

## Joining tables

#### Set up 2 tables

```{r}
# Table 1: Biological observation data
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB", "HB"),
  species = c("Trout", "Snapper", "Trout", "Cod", "Trout"),
  count = c(5, 2, 1, 3, 8)
)

# Table 2: Spatial metadata
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP", "WP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use", "Other Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12, -19.11)
)

print(observations)
print(site_metadata)
```

#### left_join(): keeps all the rows from your left table (observations) and matches data from your right table (site_metadata) wherever the key matches. If a match is missing, it fills the gaps with NA.

```{r}
# Joining metadata to our observations
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))

print(joined_data)

```

#### inner_join(): only keeps rows that have complete matches in both tables.

```{r}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))

print(matched_data)
```

#### anti_join(): filters your left table to only show rows that do not have a match in the right table (run after left join if you end up with NA values)

```{r}
# Which observations are missing from our metadata dictionary?
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))

print(missing_context)
```

## Dealing with NA Values

#### Converting legacy errors with na_if()

```{r}
 #Raw data from an old temperature logger
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code

# Convert the -999 error codes to true NA values
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))

print(fixed_logger)

```

#### Replacing NA values with coalesce()

```{r}
# Simulate some count data
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen

# Replace NA with 0 
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))

print(shark_fixed)

```

Important: coalesce() is incredibly powerful. If you have two columns, like a primary_sensor and a backup_sensor, coalesce(primary_sensor, backup_sensor) will automatically use the primary reading, but seamlessly fill in the gaps with the backup reading if the primary had failed!)

#### NaN (not a number), often happens when calculating metrics like Catch Per Unit Effort (CPUE) for a site where both the catch and the effort were zero and the function divided by zero.

```{r}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))

# Calculating CPUE (catch / effort)
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)

print(cpue_calc)

# Added line to check for NaN values
cpue_calc <- cpue_data |>
  mutate( is.nan(catch/ effort_hours), cpue = NA_real_,
    cpue = catch / effort_hours)

print(cpue_calc)

```

#### Implicit missing values: complete()

```{r}
# Important in ecology to know where species are not found, however tibbles don't typically include rows for zero catch data
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))

print(raw_catch)

# Force inclusion of hidden zero-catch data (imo)
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))

print(full_catch_matrix)
```

#### The nuclear option: drop_na(). May have to remove entire row if NA value is core response variable

```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)
print(sensor_log)

# Remove any row where salinity is missing
clean_log <- sensor_log |>
  drop_na(salinity)

print(clean_log)
```

## Example problem: Penguins, Pivots, and Relational Data

You have the clean penguins dataset, but a colleague has just emailed you a supplementary metadata file containing the GPS coordinates and the date the primary weather stations were installed on each island.

```{r}
# Make sure packages are loaded
library(tidyverse)
library(palmerpenguins)

# The messy metadata provided by your colleague
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)

print(island_metadata)


```

```{r}
# Using mutate(), stringr, and lubridate, fix the island_name column so that it perfectly matches the island column in the penguins dataset Then, convert the station_install column into a proper Date object.

clean_metadata <- island_metadata |>
  mutate(
    island_name = str_trim(island_name), # Remove leading/trailing whitespace
    island_name = str_to_title(island_name), # Convert to title case
    station_install = dmy(station_install) # Convert to Date object
  )

print(clean_metadata)
```

```{r}
# Use left_join() to attach your clean_metadata to the main penguins dataset. 
penguins_spatial <- penguins |>
  left_join(clean_metadata, by = join_by(island == island_name))

head(penguins_spatial)
```

A collaborator has asked for a clean, publication-ready table showing the maximum body mass recorded for each species, separated by island

```{r}
penguin_summary = penguins_spatial |>
  drop_na(body_mass_g) |>    #Drop missing values in the body_mass_g column
  group_by(species, island) |>   #Group by species and island
  summarise(max_mass = max(body_mass_g), .groups = "drop") |>   #Summarise data to find maximum body mass
  pivot_wider(names_from = island, values_from = max_mass)   #Pivot dataset wider so species form the rows, and islands form the columns.


print(penguin_summary)
```
